Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-13 ms/response_to_reviews.tex

We thank the editor and reviewer for their helpful comments and suggestions. We
have itemized each comment below and followed each with our reply and a list of
changes we made in response to the comment.

{\bf Responses to Senior Editor's comments:}

\begin{quote}
 I support the AE's recommendation to reject but allow resubmission. However, the bar for a resubmission will be high. My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism. Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6. In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field. If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
\end{quote}

We agree that our prior Discussion was weak and have extensively rewritten it

We argue that dry down rates are of strong ecological importance.
TODO.  Our dry down experiment occured under conduitions of very low humidity. We would expect the rates we observe in the lab to be faster than those in nature. But the differences in rates across species, the mixture effects and the resulting flammability are all important contributions of this and should represent relative differneces under field conditions.

There are very few studeis characterizing drying patterns across speceis, only one ... Kreye.

@@

\end{quote}

TODO

{\bf Responses to Associate Editor's comments:}

\begin{quote}
First, apologies to the authors for the long delay in the review process. Because of the rather contrasting assessments of the first two reviewers, I was waiting for the third reviewer to produce his/her review, but unfortunately we waited in vain. So I have to make a recommendation based on these two contrasting assessments plus my own judgement. On the one hand, in agreement with reviewer 1, I can see the novelty of this study. So far non-additivity in litter mixture flammability has been studied either at air-dry litter moisture or at fixed levels of litter moisture (e.g. Blauw et al 2015) and I do applaud the approach of bringing more field realism into these studies by allowing mixtures to dry out at different rates from expected values based on their single components. Having said that, there are also major problems with the current manuscript, as reported especially by reviewer 2 (and some additional ones by myself, see below). Besides these major problems there is a rather high degree of inaccuracy in the text.  Therefore, given the high rejection rates of often even good manuscripts to J. Ecol., I have to make a negative recommendation for this manuscript. However, given the novelty and principally nice design of the experiment, I would allow the authors to submit a strongly revised new manuscript, addressing all major and minor comments by the reviewers and myself, as a new submission. (At least one of the reviewers to check the revision would be new.) In addition to the points raised by the reviewers, I like to add some of my own:
\end{quote}

Thank you. Yes, several comments converged on the absolutely correct criticism that our discussion did not engage adequately with the literature. We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature.  We reorganized the Introduction and now do not rush the reader so quickly through the ideas.

[TODO]

@@

\begin{quote}
The authors mention leaf size explicitly as an important parameter in non-additivity (with some references including their own 2012 paper) but they do not include it in the current study. Especially at low moisture, the small individual needles of Abies should cause low ventilation and a high likelihood of negative non-additivity on flammability, as was shown by Zhao et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by reviewer 2, there is very little detail on the extent to which litter of the respective species was picked up as individual leaves, branchlets without leaves versus branchlets with leaves still attached. This is really important, also because positive versus negative non-additivity of leaf-twig mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I feel that data and/or photos indicating the size and shape of the different litter types (and if possible variability in these) are important to add and to be used in the interpretation of the results.
\end{quote}

We have added information to the methods explaining the collection in more detail. All material was naturally dropped litter which could include fine twigs and leaves. Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet and added reference to the Zhao et al 2019 results when discussing the mixture results

We have added a figure (SI fig. 1) to the SI materials showing litter particle size (containing prism volumes) distributions for each of these species. These data were collected from the stored litter. Therefore, the data show the particle size distributions of the material that was burned. We did not add these data to the main figures as they are not part of any statistical test or main conclusions, but we are happy to do so if the editor suggests.

@@

\begin{quote}
In the Methods, it was unclear how the various treatments and replicates were grouped over time, given that temperature and humidity in the fire facility could vary. Were there blocks in time with one replicate for each of the treatments (in random burning order) within a block? Or was there a different statistical design in time? 
\end{quote}

Guided by the drying rates measured in the initial trials, we staggered saturation times across species aiming to allow completely blocked design in time. Because some ignition attempts were unsuccessful, some days lacked the full complemnt of species while others later in the process sometimes had more than one replicate per species burned. We've added this information to the text [Lines X-X]

@@

\begin{quote}
Line 196 (see also reviewer 2): how can we assume that the quality of the stored litter is the same as that used in the previous study 3 years earlier, i.e. can these two datasets be safely combined?
\end{quote}

We have added detail as requested by reviewer 2. See lines X-X. The litter was initially oven dried in large paper bags, then stored in the the lab (very low humidity), for several weeks on
open racks. Then the litter was bagged and sealed and stored within garbage
bags inside cardboard boxes in a lab storage area. We saw no signs of mold or
ongoing decomposition. We minimized handling of the bags to avoid mechanical
damage and changes to little particle size.

@@

\begin{quote}
Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New Phyt reported mostly high flammability in a range of Cupressaceae.
\end{quote}

Good point and another place where we did not discuss the literature fully. We
have now added this to the discussion and made the conjecture that these
differing patterns may result from different branchelt structure across the
Cupressaceae with Australian \emph{Callitris} and the two Californian species
packing more tightly and resulting in less flammable litter than the Asian
Cupressaceae in Cornwell et al (2015).

@@

{\bf Responses to the Reviewer 1's comments}

\begin{quote}
This work compares mixtures of fuels, relative to their individual components, in terms of moisture retention and flammability. Moisture dynamics were also investigated in relation to properties of leaves of the species forming litter. This information is important in understanding flammability of surface fuels in the field.

The study has been well thought through and well executed, and the ms is generally well written. Stylistically, the main problem is that commas are missing in many sentences. This might sound trivial, but will really help the reader.
\end{quote}

We thank the reviewer for their comments and apologize for the failure in
proofreading. The reviewer should not have had to conduct copy editing and we
sincerely apologize for that. We have taken heed of the reviewers' comments and
fixed all specific grammar errors and missing symbols. For brevity, we've
omitted responses to the very simple suggested grammar fixes and simply made all
requested changes. We respond to other points individually.

@@

\begin{quote}
line 72 - the comment about natural lightning ignitions comes from nowhere, and is a  bit of a jolt for the reader - rephrase slightly, e.g. many fires result from lightning ignitions, which are concurrent with ....
\end{quote}

This portion was poorly written. We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west. See our responses to Reviewer 2, below.

@@

\begin{quote}
line 182 - affect, not effect
\end{quote}

Yes, an embarrassing error.

\begin{quote}
line 220 - see also Prior et al (2018)  Conceptualizing ecological flammability ....Fire 1, article 14
\end{quote}

Thank you. This was an oversight. We have also added more context from the
literature to our Discussion. See lines XX-XX.

\begin{quote}
lines 243/244 I was a bit puzzled here --- It would help to be more specific here --- which aspects of moisture dynamics? Which response variables did you use?  (or 'as described below')
\end{quote}

We have rewritten that section to be more precise.  See lines XX.

@@

\begin{quote}
line 403 - change to 'to the completely dry litter mixtures studied in previous work'
\end{quote}

Done. See line XX.

\begin{quote}
Figures 7 and 8 - is there a reason that you don't present the information in the same way as in Figure 6, if your main point is the discrepancy between the mixtures and the averaged individual values?
\end{quote}

We agree this does lack consistency and an early draft used the same concise format as Fig 6 format for all three figures. However, we decided that having time on the x-axis was worth the space and allows the reader more insight into the flammability effects of drying over time. Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis. There is a tension here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.

@@

{\bf Responses to the Reviewer 2's comments}

\begin{quote}

My major concerns are:
1. One motivation for the study is not valid. I do not completely agree with the authors that natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. The efficiency of individual lighting strikes in igniting a fire is affected by the moisture properties of forest fuel at the moment of lighting (Krawchuket et al. 2006, Ecology). Dry lighting that occurs without any accompanied rainfall nearby is most likely to cause wildfire because the fuel is dry enough to be ignited.
\end{quote}

This sentence was poorly written and oversimplified. We thank the reviewer
for pointing this out and we have rewritten this section. However, our badly
made point was: lighting ignitions can occur during the same general period as
precipitation. As the reviewer points out, dry strikes that find dry fuel are
more likely to lead to fires (Abatzoglou et al 2016). That the total number
lightning strikes does not well predict burned area and that ignitions occur
often from dry strikes argues for the importance of fuel moisture. Although
synoptic weather conditions that create dry and wet storms differ, the
potential for precipitation and lightning during the same overall period of
thunderstorm activity often exists (eg wet and dry convection days can occur
under overlapping conditions, Rorig and Ferguson 1999).  We've rewritten this section, see lines XX-XX.

@@

@article{van2008temporal,
  title={Temporal and spatial distribution of lightning strikes in California in relation to large-scale weather patterns},
  author={Van Wagtendonk, Jan W and Cayan, Daniel R},
  journal={Fire Ecology},
  volume={4},
  number={1},
  pages={34--56},
  year={2008},
  publisher={Springer}
}

TODO.  Get data.

\begin{quote}
2. The research question of the research is not explicitly stated both in the abstract and in the introduction. In the end of the introduction, the authors focus on specific data-collection goals, while leave the reader to figure out what the research question is. 
\end{quote}

We have edited the abstract to make the research questions clearer. Yes, our first submission jumped from stating current reserach gaps to specific methods too quickly.  We added X X

Yes, this is a good point and we have added some discussion of this.  We tried to lmit our statements to those supported by the statistical tests (Table X), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity. This, of course, obscures some details which are worth mentioning and we have edited this section in response to the reviewer's criticism. We agree that the high variability in nonadditivity across mixtures deserves highlighting. Please see our changes on lines X-X.

We have made extensive edits in response to both reviewer's suggestions.

We have changed the title.

We have rewritten much of the discussion to better include findings from the literature and rewritten the concluding paragraph.

We have editing the abstract to more clearly state the research questions reduce its length.

OK, w thought this established enough to not need citation but have added one.

We have rearranged the logic of the introduction as suggested.

We have rewritten this section and removed that phrase.

We have rewritten this for clarity. See line XX.

We were not sure what the reviewer's criticism was here but have rewritten this system to be more detailed and precise.

Good point. Fixed and in another sentence elsewhere.

We have rewritten this. See line X.

We are not sure of the criticism here.

We have broken these ideas into multiple paragraphs.

Yes, excellent point. We have added detail and citations here (lines X-X).

TODO?

@@

We have rewritten this. See line X.

\begin{quote}
Line 136: delete “on”. Revise “ecologically-relevant”.
\end{quote}

TODO

We assume that the reviewer's complaint is that ``ecologically-relevant'' is vague.

@@

\begin{quote}
Line 141: revise “(36 36N, 118 42 W)”.
\end{quote}

Symbols fixed throughout.

\begin{quote}
Line 147: at least for Quercus, leaves fallen in the previous year will be partly decomposed in the summer.
\end{quote}

Yes. We have added some detail to this and also now provide particle size distributions for each species measured on the actual stored litter (SI fig 1).

@@

\begin{quote}
Line 155: add more details about how the litter was collected?
\end{quote}

We have added information to the methods explaining the collection in more detail. See lines XX and our response to the Associate Editor, above.

\begin{quote}
Line 157: is it possible that all litter has less than 5\% fuel moisture when air-dried?
\end{quote}

We have corrected the methods because the litter was initially oven dried in large paper bags, then stored on racks in the low humidity lab. Our subsample measurements showed less than 5\% moisture.  See lines X-X.

@@

\begin{quote}
Line 158: the collecting of the leaf litter is in 2012. The conduction of the experiment is in 2015. Can the author explain how do they save the litter?
\end{quote}

We have added detail on the long term dry storage. See lines X-X.

@@

\begin{quote}
Line 185: I do not think this sentence is necessary because whether or not petioles should be included in SLA measurements depends on the research question at hand.
\end{quote}

Agreed, fixed.

\begin{quote}
Line 201: the fuel depth was controlled or not?
\end{quote}

No, mass was fixed, width and length of the burning tray was fixed (15\,cm by 1.5\,m), and depth was allowed to vary.  We edited this for clarity. See lines X-X

@@

\begin{quote}
Line 216: revise “mass loss” to “fuel consumption”.
\end{quote}

Fixed.

\begin{quote}
Line 224: if not all flammability data was used in the analysis, then why measure them?
\end{quote}

Flammability is a multivariate trait. To avoid collinearity and to simplify
models, two general approaches are possible: either reduce and transform the
response variables and create synthetic variables through a method such as PCA,
or pick a few variables representative of the axes of variation. When the
experiment was designed, we had intended to use the first method. The loss of
several days of thermocouple data made that an unattractive and unnecessary
option, so we opted for the simpler to explain method of picking two traits
already known to correlate with different axes of litter flammability. I envy
the scientist who has never dealt with unforeseen events.

@@

\begin{quote}
Line 242: revise “one leaf trait and one litter trait” to “one litter particle trait and one litter bed traits”.
\end{quote}

TODO

\begin{quote}
Line 307: revise “at that point”.
\end{quote}

Although we are unsure of the criticism here, we have rewritten the sentence.

@@

\begin{quote}
Lines 314-320: this indicates that even in the same genus, species vary in flammability.
\end{quote}

Yes.  This is not surprising in the case of these three pines where one has quite different needles. But our \emph{a priori} groups were based on genus.

@@

\begin{quote}
Line 401: “striking”？
\end{quote}

We removed the offending word.

\begin{quote}
Line 410: “not met by the constituent species alone”?
\end{quote}

We don't understand the criticism. Perhaps it is the word ``alone'' that is ambiguous?  We changed ``alone'' to ``in monoculture'' although we thought the original wording was clear.

@@

\begin{quote}
Line 415: “demonstrates”?
\end{quote}

We don't understand the criticism. Academic writing commonly uses present tense to describe results that still apply. There is no grammatical error.

### 2019-12-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Progress towards R1

      \item To fill these gaps in knowledge, determined
    moisture absorption and retention characteristics of the litter from
    multiple tree species, investigated what litter traits influenced these
    processes, and tested for nonadditive moisture dynamics and nonadditive
    flammability in multi-species mixtures. The litter from eight tree species of a
    mixed-conifer forest was saturated and allowed to dry to determine moisture
    absorption capacity and drying rates. Burn trials were performed in moist
    litter beds of single species and of mixtures to establish flammability
    response to the dry-down process.

@@

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarf+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarf+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018}.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+Cornwell+etal-2016}. It is possible that
similar nonadditive effects may influence litter moisture content directly and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}.

@@

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
the influence of moisture on the flammability across these species.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. Here, we examined if this positive nonadditivity was
maintained in mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are
nonadditive effects as observed in oven-dried litter. We tested if the
nonadditive effects on reported in previous studies occur under [TODO CHECK]
ecologically-relevant moisture levels, and whether fuel moisture effects
strengthen or moderate such nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed or lightly decomposed). Leaf litter was
collected in the summer (mid-June to mid-July) in 2012 from 21 separate sites
across the study area. We collected from a minimum of four populations per
species with each population 9--32\,km distant from its neighbors. An exception
to this was \emph{A. magnifica} which is restricted to the higher elevations
and could only be collected from one area due to designated wilderness
boundaries and logistical constraints. Within that area, \emph{A. magnifica}
litter was collected from over 10 different individuals. For the remaining
species, the collection involved 2--4 individual trees at each site which were
separated from one another by at least 10\,m. The litter was collected 2\,m
away from the trunk to obtain a more uniform sample of leaf and fine branchlet
litter, because bark and heavier branches tend to fall closer to the tree
trunk. The top 2.5-5\,cm of litter were collected by raking and shoveling into
large bags. All litter was air dried to $< 5$\% fuel moisture by dry weight,
then weighed and divided into moisture trials. The collected litter was dried
in large batches, then stored in sealed plastic bags inside cardboard boxes in
a lab storage space. We minimized handling and moving the bagged litter to
avoid mechanical damage and changes to the particle size distributions. The
moisture manipulations and subsequent burning trials were conducted in
April--August 2015.

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption (specific
leaf area), and one litter trait known to directly affect flammability and
likely to influence drying rate and litter packing density (litter bulk
density). [TODO check] Specific leaf area was measured on freshly fallen litter
by sampling 5 individuals per species. This measurement included petiole and
for the scale-leaved species included the full excised branchlet. Therefore,
this is not directly comparable to typical specific leaf area measurements on
fresh leaves. Litter bulk density was obtained in this experiment for each
basket by dividing the dry mass (450\,g) by the estimated litter volume (basket
width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. [TODO give depths] This design allows fire to reach
constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal
rulers equally spaced along the apparatus allowed visual determination of
maximum flame height. Litter was ignited with a propane torch; maximum
adiabatic flame temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. A timer was used to record total
time to ignition and duration of flaming combustion. The flammability trials
followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and fuel consumption were recorded, but flame spread rate and flame
height were recorded as zero. Our thermocouple system failed or data was lost
for two days of burning trials, therefore we have omitted those data from this
analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and
percent fuel consumed captures a portion of the second major axis of
flammability variation. Our analysis, therefore, focused on these two
relatively orthogonal measures of flammability to avoid examining multiple
redundant dependent variables.

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
between 10 am and 3 pm.

@@

Dry-down curves were created based on the drying experiment, fitted with an
exponential decay curve to each species.:

@@

Across these eight species specific leaf area had a positive association with maximum
water retention (SI Fig 4, linear model $P$ = 0.003) and litter
bulk density had a negative effect on desiccation rate (SI Fig 5,
linear model $P$ = 0.029). ANOVA results are shown in Tables
\ref{tab:mc_di_anova}.

@@

Based on visual inspection of the dry down curves and the estimated marginal
means for maximum water retainability and desiccation rate (SI Figs 2--3), we
selected four species to represent the full range of litter moisture dynamics
for use in the litter mixtures and examined all four three-species combinations
of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus
  decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.

>>>>

      \item To fill these gaps in knowledge, we determined
    moisture absorption and retention characteristics of the litter from
    multiple tree species, investigated what litter traits influenced these
    processes, and tested for nonadditive moisture dynamics and nonadditive
    flammability in multi-species mixtures. The litter from eight tree species of a
    mixed-conifer forest was saturated and allowed to dry to determine moisture
    absorption capacity and drying rates. Burn trials were performed in moist
    litter beds of single species and of mixtures to establish flammability
    response to the dry-down process.

@@

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible that
similar nonadditive effects may influence litter moisture content directly and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}.

@@

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
the influence of moisture on the flammability across these species.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. Here, we examined if this positive nonadditivity was
maintained in mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are
nonadditive effects as observed in oven-dried litter. We tested if the
nonadditive effects on reported in previous studies occur under [TODO CHECK]
higher moisture levels, and whether fuel moisture effects
strengthen or moderate such nonadditivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed or lightly decomposed). Leaf litter was
collected in the summer (mid-June to mid-July) in 2012 from 21 separate sites
across the study area. We collected from a minimum of four populations per
species with each population 9--32\,km distant from its neighbors. An exception
to this was \emph{A. magnifica} which is restricted to the higher elevations
and could only be collected from one area due to designated wilderness
boundaries and logistical constraints. Within that area, \emph{A. magnifica}
litter was collected from over 10 different individuals. For the remaining
species, the collection involved 2--4 individual trees at each site which were
separated from one another by at least 10\,m. The litter was collected 2\,m
away from the trunk to obtain a more uniform sample of leaf and fine branchlet
litter, because bark and heavier branches tend to fall closer to the tree
trunk. The top 2.5-5\,cm of litter were collected by raking and shoveling into
large bags. All litter was oven dried to $< 5$\% fuel moisture by dry weight,
then weighed and divided into moisture trials. The collected litter was stored
in large paper bags and cardboard boxes in a lab storage space. We minimized
handling and moving the bagged litter to avoid mechanical damage and changes to
the particle size distributions. The moisture manipulations and subsequent
burning trials were conducted in April--August 2015.

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption (specific
leaf area), and one litter trait known to directly affect flammability and
likely to influence drying rate and litter packing density (litter bulk
density). Specific leaf area was measured on freshly fallen litter
by sampling 5 individuals per species. This measurement included petiole and
for the scale-leaved species included the full excised branchlet. Therefore,
this is not directly comparable to typical specific leaf area measurements on
fresh leaves. Litter bulk density was obtained in this experiment for each
basket by dividing the dry mass (450\,g) by the estimated litter volume (basket
width x length x litter depth). To examine litter particle size distributions in the stored litter, we subsampled 10 approximately 2\,g samples per species from the stored litter. Each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. Therefore, fuel mass and burning tray length and width
were fixed, but fuel depth was allowed to vary. We recorded the actual fuel
depth to calculate litter bed density before burning. This design allows fire
to reach constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two
graduated metal rulers equally spaced along the apparatus allowed visual
determination of maximum flame height. Litter was ignited with a propane torch;
maximum adiabatic flame temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. A timer was used to record total
time to ignition and duration of flaming combustion. The flammability trials
followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and fuel consumption were recorded, but flame spread rate and flame
height were recorded as zero. Our thermocouple system failed or data was lost
for two days of burning trials, therefore we have omitted those data from this
analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018} and
percent fuel consumed captures a portion of the second major axis of
flammability variation. Our analysis, therefore, focused on these two
relatively orthogonal measures of flammability to avoid examining multiple
redundant dependent variables.

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
between 10 am and 3 pm. We scheduled burns in a randomized blocked design
aiming to burn one replicate of each species or mixture on single burn day.
This required scheduling species saturation and during start based on the dry
down curves.

@@

Dry-down curves were created based on the drying experiment, fitted with an
exponential decay curve to each species:

@@

Across these eight species specific leaf area had a positive association with maximum
water retention (SI Fig 5, linear model $P$ = 0.003) and litter
bulk density had a negative effect on desiccation rate (SI Fig 6,
linear model $P$ = 0.029). ANOVA results are shown in Tables
\ref{tab:mc_di_anova}.

@@

Based on visual inspection of the dry down curves and the estimated marginal
means for maximum water retainability and desiccation rate (SI Figs 3--4), we
selected four species to represent the full range of litter moisture dynamics
for use in the litter mixtures and examined all four three-species combinations
of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus
  decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.

Mixture dry down rates and maximum water retention showed some relationship to
the dry down curves for individual constituent species (see SI Fig. 7). For
example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
Fig. 7) has the lowest initial retention. Initial water retention and drying
rates differed across mixtures (linear model based on log of moisture content,
SI Tables 8 and 9).
